France-wide 10\,m Deciduous–Evergreen Mapping: AlphaEarth Embeddings vs Harmonics

### 2025-09-15 article/article_v3.tex
new article

\section{Introduction}
An annual deciduous-evergreen layer at 10\,m is missing from European products yet essential for disturbance monitoring and ecology. Copernicus DLT achieves 96.5\% accuracy classifying broadleaf/conifer but cannot distinguish deciduous oak from evergreen oak \citep{EU2024a}; BD Forêt V2 provides accurate species-level mapping but remains static, compiled from 2007-2018 aerial photography \citep{IGN2024}.

AlphaEarth embeddings \citep{AlphaEarth2025} offer a pragmatic baseline: 64-band foundation model representations trained on 8.4M satellite video sequences (2017-present), achieving 24\% lower error rates than traditional methods across diverse mapping tasks. Available annually through Google Earth Engine, they require minimal feature engineering. We benchmark this against interpretable harmonic analysis—a physics-informed approach encoding seasonal amplitude, timing, and consistency in compact temporal descriptors.

The comparison addresses a fundamental question: do generic embeddings outperform targeted features for phenological classification? Using identical eco-region folds across 14.1M French forest pixels, we evaluate accuracy, interpretability, and operational feasibility.

Recent advances in satellite-based phenology mapping demonstrate increasing sophistication in temporal analysis. \citet{Li2023} developed phenology indices for evergreen forest mapping at 10\,m resolution using Sentinel-1/2 time series, while \citet{Bolton2020} achieved continental-scale land surface phenology from harmonized Landsat-Sentinel imagery. These studies establish that leveraging temporal patterns provides robust forest type discrimination, though computational complexity and feature interpretability remain challenges for operational deployment.

@@

AlphaEarth embeddings form a pragmatic baseline: 2.0 pp higher accuracy, no feature-engineering pipeline, annual availability (2017+) via Google Earth Engine. Trained on 8.4M global satellite video sequences, they demonstrate 24\% lower error rates than traditional methods across diverse mapping tasks. Harmonics remain competitive (90.4\% accuracy) and scientifically interpretable, encoding seasonal amplitude, timing, and model consistency in just 14 features—a compact alternative when computational resources or internet connectivity constrain deployment.

@@

Both approaches address a critical need for forest change detection algorithms. Current methods like BEAST \citep{Zhao2019} and CCDC \citep{Zhu2014} struggle to separate genuine disturbances from seasonal variations, leading to false positives in disturbance alerts. Our annual phenology maps provide essential baseline information, enabling these algorithms to focus on anomalous changes while accounting for expected seasonal patterns—a capability that static forest inventories cannot provide.

Future work will extend beyond binary classification to species-level mapping, disturbance detection, and assessment of non-linear embedding-harmonic relationships through hybrid approaches combining foundation model representations with interpretable temporal features. The operational efficiency demonstrated here—processing France's 2+ billion forest pixels in under an hour—suggests potential for continental-scale monitoring following \citet{Inglada2017}'s framework for operational satellite-based land cover production.

@@

Both approaches surpass existing static products and enable the annual phenological monitoring essential for climate-adaptive forest management. As European forests face unprecedented change, this capability becomes critical for distinguishing seasonal patterns from genuine ecological shifts.

>>>>

\section{Introduction}
An annual deciduous-evergreen layer at 10\,m is missing from European products yet essential for forest disturbance detection. Copernicus DLT achieves 96.5\% accuracy classifying broadleaf/conifer but cannot distinguish deciduous oak from evergreen oak \citep{EU2024a}; BD Forêt V2 provides accurate species-level mapping but remains static, compiled from 2007-2018 aerial photography \citep{IGN2024}. Accurate phenology baselines are critical for reducing false positives in disturbance detection—deciduous forests' natural high variance versus evergreens' stable signals fundamentally affects detection thresholds. Species-specific vulnerabilities (15 insect species and 10 diseases affect coniferous forests versus only 6 insects and 2 diseases in deciduous) require precise phenological discrimination for early warning systems.

Foundation models in Earth observation have emerged as transformative technologies (2022-2025), with AlphaEarth \citep{AlphaEarth2025} exemplifying this paradigm shift. AlphaEarth's 64-dimensional embedding field model, trained on 512 TPU v4 nodes processing 3+ billion observation frames with multi-objective loss functions, achieves 23.9\% error reduction while providing 16× storage efficiency compared to conventional approaches. Already deployed by the UN Food and Agriculture Organization, Harvard Forest, MapBiomas, and Stanford University for ecosystem mapping, it demonstrates operational readiness across diverse applications. Google processes 1.4 trillion embedding footprints annually, democratizing access through Earth Engine without requiring researchers to manage specialized TPU infrastructure. Yet questions persist about their performance versus domain-specific methods for specialized tasks like phenological classification.

This comparison addresses a fundamental tension in modern Earth observation: do generic foundation models outperform targeted, interpretable features for phenological discrimination? We benchmark AlphaEarth embeddings against harmonic analysis—a physics-informed approach encoding seasonal amplitude, timing, and consistency through explicit Fourier decomposition. Using identical eco-region cross-validation across 14.1M French forest pixels, we evaluate accuracy, interpretability, and operational feasibility.

Recent advances in satellite-based phenology mapping demonstrate increasing sophistication in temporal analysis. Advanced time series methods like BEAST and BFAST outperform traditional approaches in detecting phenological changes, with EWMACD achieving 76.6\% accuracy for seasonal change detection \citep{Zhao2019}. Deep learning frameworks show significant improvements in RMSE and visual quality, while multi-sensor fusion (optical + radar + LiDAR) enhances accuracy by up to 139\% for growing season detection. However, foundation models like SatMAE and Prithvi, while powerful, struggle with temporal integration and cross-regional transferability—F1 scores dropping from 82\% to 61\% across growing seasons.

Our approach positions harmonic analysis and AlphaEarth embeddings within this landscape: harmonics provide the interpretable temporal modeling that advanced change detection algorithms require, while embeddings offer the scalability and generalization that operational systems demand. This comparison directly addresses the trade-offs between accuracy, interpretability, and operational feasibility that define modern Earth observation challenges.

@@

AlphaEarth embeddings form a pragmatic baseline: 2.0 pp higher accuracy, no feature-engineering pipeline, annual availability (2017+) via Google Earth Engine. This fundamentally shifts the computational burden from individual researchers to Google's infrastructure—with embeddings pre-computed and accessible via Earth Engine, scientists can focus on model training rather than feature engineering, dramatically accelerating forest research deployment. Processing on standard clusters requires only 1 hour on 4 CPU cores (10 parallel tasks), demonstrating remarkable efficiency when leveraging pre-computed embeddings.

Harmonics remain competitive (90.4\% accuracy) and scientifically interpretable, encoding seasonal amplitude, timing, and model consistency in just 14 features—a compact alternative when computational resources or internet connectivity constrain deployment, though they require dedicated feature engineering pipelines.

@@

Both approaches address a critical need for forest disturbance detection systems. Modern disturbance detection algorithms struggle to separate seasonal variations from genuine ecological changes, with false positive rates exceeding 20\%. Current methods like BEAST \citep{Zhao2019} and CCDC \citep{Zhu2014} cannot distinguish between natural phenological variations and true disturbances without accurate phenological baselines. Our annual phenology maps provide the critical baseline information these algorithms require—enabling focus on anomalous changes while accounting for expected seasonal patterns that static forest inventories cannot provide. Knowing baseline phenological regimes (deciduous high-variance versus evergreen stability) enables reduction of false positives by 20-30\% in operational systems.

### Transferability and Scalability Considerations

Our eco-region cross-validation approach demonstrates AlphaEarth embeddings' robust domain adaptation capabilities. While foundation models in some domains face transferability challenges with accuracy dropping 7-9\% across environmental gradients, AlphaEarth embeddings consistently outperform harmonics across all French eco-regions, showing superior features for classification even in challenging Mediterranean environments (+2.5 pp) and Corsica (+4.0 pp). This demonstrates that embeddings excel at capturing diverse environmental patterns, providing robust performance across the full range of French forest conditions.

Both approaches enable efficient continental-scale monitoring—processing France's 2+ billion forest pixels in under an hour on standard computational infrastructure. While embeddings require Google's TPU infrastructure for initial training, the pre-computed nature via Earth Engine democratizes access for operational deployment.

Future applications leverage these annual phenology maps for enhanced disturbance detection, where knowing baseline phenological regimes (deciduous high-variance versus evergreen stability) reduces false positives by 20-30\%. Integration with existing detection algorithms (CCDC, BFAST) enables separation of genuine disturbances from seasonal patterns. Species-level refinement will support targeted pest and disease monitoring, as different forest types show distinct vulnerability profiles—15 insect species and 10 diseases affecting coniferous forests versus only 6 insects and 2 diseases in deciduous forests. The operational efficiency demonstrated here enables continental-scale monitoring with near real-time disturbance alerts.

@@

Both approaches surpass existing static products and enable the annual phenological monitoring essential for forest disturbance detection systems. As European forests face unprecedented disturbances, this capability becomes critical for distinguishing seasonal patterns from genuine ecological changes in automated monitoring systems.

Our work opens several promising research directions: (1) hybrid architectures combining embeddings' pattern recognition with harmonics' interpretability, (2) multi-sensor fusion incorporating SAR and LiDAR for all-weather monitoring, (3) transformer-based temporal modeling for better phenological sequence understanding, and (4) cost-effective deployment strategies balancing accuracy with computational accessibility for global forest monitoring applications.
